---
title: "MB5370 Module 04. Workshop 1 - Introduction"
author: "Alexander Barkley"
output: html_document
date: "2024-03-22"
---
# --------@
# Module 4 - Data in R - Workshop 1
# Alexander Barkley
# 19-03-2025
# --------@

```{r}
# install.packages("usethis")
# usethis::use_github()
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Install and load tidyverse packages 
```{r}
# install.packages("tidyverse") 
library("tidyverse")
```


### Load the data (mpg is built into ggplot2) 
```{r}
data(mpg)
```


### Quick data checks 
```{r}
head(mpg)
glimpse(mpg)
summary(mpg)
```


### Creating my first ggplot
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```

The plot shows a negative relationship between engine size (displ) and fuel efficiency (hwy). In other words, cars with big engines use more fuel. What does this say about fuel efficiency and engine size?

The plot shows a negative correlation between engine size (displ) and fuel efficiency (hwy), meaning larger engines consume more fuel and have lower highway MPG. This highlights the trade-off between power and efficiency—smaller engines are more fuel-efficient, while larger engines provide more power at the cost of fuel economy.

### Understanding grammar of graphics 
What happens if we just try to run ggplot on its own?
```{r}
ggplot() 
```

### Or with just the data specified? 
```{r}
ggplot(data = mpg)
```

We need to map some aesthetics!!

### When you’re creating a plot, you essentially need two attributes of a plot: a geom and aesthetics. 
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, colour = class))
```


### Change point colour by class:
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, colour = class))
```

### Change point size by class:
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, size = class))
```
Note the warning!!!


### Change transparency (alpha) by class
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, alpha = class))
```
Another warning!!
Question: When would using alpha (or size be appropriate??)

### Change point shape by class:
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, shape = class))
```


### Make all points blue
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), color = "blue")
```


### Faceting

### facet_wrap
Split the plot out by car type (class)
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_wrap(~ class, nrow = 2)
```


### facet_grid
A separate facet for each combination of drive-type (e.g. 4WD) * number of cylinders
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(drv ~ cyl)
```
Note that there are no occurrences of 5 cylinder 4WDs OR RWD vehicles!


### facet_grid by just row (or column)
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(cyl ~ .)
  #facet_grid(. ~ cyl) # Alternatively

```

### Exercise:
Read ?facet_wrap. What does nrow do? What does ncol do? What other options control the layout of the individual panels? 

```{r}
?facet_wrap
ncol
```


### Lines
We already used points:
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```


However, ggplot2 can use a variety of geom objects to represent the data. Here, we might want to use bar plots, line charts, boxplots and so on. Well we can handle this issue in ggplot directly using a different geom to plot the same data. Here, instead of plotting points, we will use a smooth line. 

### To display the same data as a smooth line fit through the points use geom_smooth().
```{r}
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy))
```
### Question: whey don't we use geom_line() here? What would that look like? ###


So let’s recap. A geom is an object that your plot uses to represent the data. To change the geom type in your plot, simply change the geom function that you add to your plot template. Sometimes you may want to try a few things out, in which case you could use comments to help you remember what worked and what didn’t. 

### Using comments (#)
```{r}
ggplot(data = mpg) + 
#  geom_point(mapping = aes(x = displ, y = hwy)) # points horrible 
  geom_smooth(mapping = aes(x = displ, y = hwy)) # try smooth line
```

Question: how does R work out where to plot the line??? Can use the chunk output to inform us. Can also use the help menu.

### Changing linetype
```{r}
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy, linetype = drv)) # Can also use "lty = " for the same result
```


### Grouping
```{r}
ggplot(data = mpg) +
  geom_smooth(mapping = aes(x = displ, y = hwy, group = drv))
```
 
### Change line colour based on drv value
```{r}
ggplot(data = mpg) +
  geom_smooth(
    mapping = aes(x = displ, y = hwy, color = drv),
    show.legend = FALSE,
  )
```

### Multiple geoms
We already did this one way
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  geom_smooth(mapping = aes(x = displ, y = hwy))
```

A better way...
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth()
```


### Can still manipulate each geom/layer separately:
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) + 
  geom_smooth()
```

Now we have different colours to indicate the raw data (geom_point) and one colour for the overall relationship (geom_smooth). Neat!


```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) + 
  geom_smooth(data = filter(mpg, class == "subcompact"), se = FALSE)
```
Question: what does the "se = FALSE" do?


# Exercise: MAKEEEEE SURRREEE TOOOO DO IITTTTT
1.  What geom would you use to draw a line chart? A boxplot? A histogram? An area chart?
2.  Run this code in your head and predict what the output will look like. Then, run the code in R and check your predictions.

```{r}

```


3.  Will these two graphs look different? Why/why not?
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth()

ggplot() + 
  geom_point(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_smooth(data = mpg, mapping = aes(x = displ, y = hwy))
```


# Transformations and stats

## Load data
```{r}
data("diamonds")
glimpse(diamonds)
```

# Bar plot
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut))
```

# Overriding defaults
Make some new data
```{r}
demo <- tribble(
  ~cut,         ~freq,
  "Fair",       1610,
  "Good",       4906,
  "Very Good",  12082,
  "Premium",    13791,
  "Ideal",      21551
)
demo
```


```{r}
ggplot(data = demo) +
  geom_bar(mapping = aes(x = cut, y = freq), stat = "identity")
```
Question: Why did we specify now specify a y axis?

# What if we want to know the proportion rather than the absolute frequency?
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, y = stat(prop), group = 1))
```

Question: does anyone get the warning that's in the workbook?
Warning: `stat(prop)` was deprecated in ggplot2 3.4.0.
If so, can use # geom_bar(mapping = aes(x = cut, y = stage(after_stat = prop), group = 1))

# Plotting statistical details
```{r}
ggplot(data = diamonds) + 
  stat_summary(
    mapping = aes(x = cut, y = depth),
    fun.min = min,
    fun.max = max,
    fun = median
  )
```


# Aesthetic adjustments adjustments
Another way to boost the way you can convey information with plots using ggplot2 is to use aesthetics like colour or fill to change aspects of bar colours. We already did this once, but there are multiple options available to you, including changing the fill our outline colours.

```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, colour = cut))
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = cut))
```
Question: Does anyone notice anything different in the colour scheme of this plot?
(Hint: It's in the viridis colour palette (colour blind friendly), but why is it different from the colour palette we used earlier?)
Check out the difference:
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) 

ggplot(data = diamonds, mapping = aes(x = color, y = depth)) + 
  geom_point(mapping = aes(color = cut)) 
```


# Filling by a variable
Now try using these aesthetics to colour by another variable like clarity. Notice how the stacking is done automatically. This is done behind the scenes with a position argument.
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity))
```


# To alter transparency (alpha)
```{r}
ggplot(data = diamonds, mapping = aes(x = cut, fill = clarity)) + 
  geom_bar(alpha = 1/5, position = "identity")
```


# To colour the bar outlines with no fill colour
```{r}
ggplot(data = diamonds, mapping = aes(x = cut, colour = clarity)) + 
  geom_bar(fill = NA, position = "identity")
```


# Position adjustments
position = "fill" works like stacking, but makes each set of stacked bars the same height.
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "fill")
```


# position = "dodge" 
Places overlapping objects directly beside one another.
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "dodge")
```

# Jittering
position = "jitter" adds a small amount of random noise to each point to avoid overplotting when points overlap. This is useful for scatterplots but not barplots.
```{r}
ggplot(data = mpg) + 
  #geom_point(mapping = aes(x = displ, y = hwy), position = "jitter")
  geom_point(mapping = aes(x = displ, y = hwy))
```

# Assignment: DOUBLEEEEEEE CHECCKKKKKKKK

# Plot Deconstruction
```{r}
# Load necessary libraries
library(ggplot2)
library(dplyr)

# Loading the dataset 
data <- read.csv('/Users/alexanderbarkley/Downloads/assignment 1 mod 4 dataset.csv', skip = 1)  # Skip first row

# Rename columns for clarity
colnames(data) <- c("Distance_Rainy", "Semivariance_Rainy", "Distance_Dry", "Semivariance_Dry")

# Convertimg columns to numeric
df <- data %>%
  mutate(across(everything(), as.numeric))

# Convert data to long format for ggplot**
df_long <- bind_rows(
  data.frame(Distance = df$Distance_Rainy, Semivariance = df$Semivariance_Rainy, Season = "Rainy"),
  data.frame(Distance = df$Distance_Dry, Semivariance = df$Semivariance_Dry, Season = "Dry")
) %>%
  filter(!is.na(Distance) & !is.na(Semivariance)) %>%  # Remove missing values
  arrange(Season, Distance)  # Ensure correct order for smooth lines

# Create the plot
ggplot(df_long, aes(x = Distance, y = Semivariance, color = Season)) +
  geom_point(size = 2, alpha = 1, ) +  # Scatter points
  geom_smooth(method = "loess", span = 0.5, se = 0) +  # Explicitly use LOESS, using non-parametric regression method used to smooth data trends in scatter plots 
  theme_minimal() +
  labs(title = "Fish species richness for rainy and dry seasons in 1993–1994",
       x = "Distance (km)",
       y = "Semivariance",
       color = "Season") +  # Legend title
  scale_color_manual(values = c("Rainy" = "lightblue", "Dry" = "lightcoral"))  

```




# --------@
# Module 4 - Data in R - Workshop 2
# Alexander Barkley
# 20-03-2025
# --------@

# Labels
```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se.e = FALSE) +
  labs(title = "Fuel efficiency generally decreases with engine size")
# note: "[38;5;252m`geom_smooth()` using method = 'loess' and formula = 'y ~ x'[39m" - this states that it doesnt have data for some part of thw curve therefore it just figures or estimates where it would be to make the smooth line.
```
# Adding subtitles and caption
```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) + # "$class - unexpected character here)
  geom_smooth(se = FALSE) +
  labs(
    title = "Fuel efficiency generally decreases with engine size",
    subtitle = "Two seaters (sports cars) are an exception because of their light weight",
    caption = "Data from fueleconomy.gov"
  )

```

# Using "labs" to replace axis titles and labels; instead of using aesthetics, we can specify what we want our x and y axes to be.
```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) + # "classs" error adjusted
  geom_smooth(se = FALSE) +
  labs(
    x = "Engine displacement (L)",
    y = "Highway fuel economy (mpg)",
    colour = "Car type"
  )
```

### 3.2 - Annotations
```{r}
best_in_class <- mpg %>%
  group_by(class) %>%
  filter(row_number(desc(hwy)) == 1)

ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) +
  geom_text(aes(label = model), data = best_in_class)
```

### 3.3 - Adding scales to graph

```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class))
```
Doing this just ensures ggplot aadss scales in the background, however, scales can be specified.

```{r}
?scale_x_continuous()

limits = c(0,7)

ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) +
  scale_x_continuous() +
  scale_y_continuous() +
  scale_colour_discrete()
```

### 3.4 - Axis ticks

```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  scale_y_continuous(breaks = seq(15, 40, by = 5))

```

```{r}
# seq(15, 40, by = 5 ?? 

ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  scale_x_continuous(labels = NULL) +
  scale_y_continuous(labels = NULL) # adds the points on the axis but no labels
```


### 3.5 - Legends and colour schemes

```{r}
base <- ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class))

base + theme(legend.position = "left")
```

```{r}
base + theme(legend.position = "top")
```

```{r}
base + theme(legend.position = "bottom")
```

```{r}
base + theme(legend.position = "right")
```

```{r}
base + theme(legend.position = "none")
```

### 3.6 - Replacing a scale

```{r}
ggplot(diamonds, aes(carat, price)) +
  geom_bin2d() + 
  scale_x_log10() + 
  scale_y_log10()
```

```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = drv))

ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = drv)) +
  scale_colour_brewer(palette = "Set1")
```

```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = drv, shape = drv)) +
  scale_colour_brewer(palette = "Set1")
```

```{r}
presidential %>%
  mutate(id = 33 + row_number()) %>%
  ggplot(aes(start, id, colour = party)) +
    geom_point() +
    geom_segment(aes(xend = end, yend = id)) +
    scale_colour_manual(values = c(Republican = "red", Democratic = "blue"))
```


```{r}
#install.packages('viridis')
#install.packages('hexbin')
library(viridis)
library(hexbin)
```

```{r}
?tibble

df <- tibble( # note we're just making a fake dataset so we can plot it
  x = rnorm(10000),
  y = rnorm(10000)
)
ggplot(df, aes(x, y)) +
  geom_hex() + # a new geom!
  coord_fixed()

ggplot(df, aes(x, y)) +
  geom_hex() +
  viridis::scale_fill_viridis() +
  coord_fixed()

```
### 3.7 - Themes
```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  theme_bw()
```

```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  theme_light()
```

```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  theme_classic()
```


```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  theme_dark()
```

### Developing your own theme
```{r}
theme (panel.border = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        legend.position="right",
        legend.title=element_blank(),
        legend.text=element_text(size=8),
        panel.grid.major = element_blank(),
        legend.key = element_blank(),
        legend.background = element_blank(),
        axis.text.y=element_text(colour="black"),
        axis.text.x=element_text(colour="black"),
        text=element_text(family="Times New Roman"))
```

### Saving and exporting plots

```{r}
ggplot(mpg, aes(displ, hwy)) + geom_point()

ggsave("my-plot.pdf")
#> Saving 7 x 4.32 in image
```












### Assignment - QFISH

```{r}
shark <- read.csv('/Users/alexanderbarkley/Downloads/Number caught by area, calendar year and species group (QFISH).csv')
view(shark)
```










# --------@
# Module 4 - Data in R - Workshop 3
# Alexander Barkley
# 09-04-2025
# --------@


### Data Wrangling in R

```{r}
library(tidyverse)
```

### Exercise
```{r}
For each of the sample tables, describe what each observation and each column represents.

table1
#> # A tibble: 6 × 4
#>   country      year  cases population
#>   <chr>       <int>  <int>      <int>
#> 1 Afghanistan  1999    745   19987071
#> 2 Afghanistan  2000   2666   20595360
#> 3 Brazil       1999  37737  172006362
#> 4 Brazil       2000  80488  174504898
#> 5 China        1999 212258 1272915272
#> 6 China        2000 213766 1280428583


table2
#> # A tibble: 12 × 4
#>   country      year type           count
#>   <chr>       <int> <chr>          <int>
#> 1 Afghanistan  1999 cases            745
#> 2 Afghanistan  1999 population  19987071
#> 3 Afghanistan  2000 cases           2666
#> 4 Afghanistan  2000 population  20595360
#> 5 Brazil       1999 cases          37737
#> 6 Brazil       1999 population 172006362
#> # ... with 6 more rows


table3
#> # A tibble: 6 × 3
#>   country      year rate 
#> * <chr>       <int> <chr>            
#> 1 Afghanistan  1999 745/19987071     
#> 2 Afghanistan  2000 2666/20595360    
#> 3 Brazil       1999 37737/172006362  
#> 4 Brazil       2000 80488/174504898  
#> 5 China        1999 212258/1272915272
#> 6 China        2000 213766/1280428583

```


### Sketch out the processes you would use to calculate the rate for table2 and table3. You will need to perform four operations:
### Extract the number of TB cases per country per year
### Extract the matching population per country per year
### Divide cases by population, and multiply by 10,000
### Store back in the appropriate place

```{r}
### Example 

# Compute rate per 10,000
table1 %>% 
  mutate(rate = cases / population * 10000)
#> # A tibble: 6 × 5
#>   country      year  cases population  rate
#>   <chr>       <int>  <int>      <int> <dbl>
#> 1 Afghanistan  1999    745   19987071 0.373
#> 2 Afghanistan  2000   2666   20595360 1.29 
#> 3 Brazil       1999  37737  172006362 2.19 
#> 4 Brazil       2000  80488  174504898 5.61 
#> 5 China        1999 212258 1272915272 1.67 
#> 6 China        2000 213766 1280428583 1.67

# Compute cases per year
table1 %>% 
  count(year, wt = cases)
#> # A tibble: 2 × 2
#>    year      n
#>   <int>  <int>
#> 1  1999 250740
#> 2  2000 296920

# Visualise changes over time
library(ggplot2)
ggplot(table1, aes(year, cases)) + 
  geom_line(aes(group = country), colour = "grey50") + 
  geom_point(aes(colour = country))

### Exercise answer

table2
#> # A tibble: 12 × 4
#>   country      year type           count
#>   <chr>       <int> <chr>          <int>
#> 1 Afghanistan  1999 cases            745
#> 2 Afghanistan  1999 population  19987071
#> 3 Afghanistan  2000 cases           2666
#> 4 Afghanistan  2000 population  20595360
#> 5 Brazil       1999 cases          37737
#> 6 Brazil       1999 population 172006362
#> # ... with 6 more rows
# Compute rate per 10,000 for table two

### 1. Firstly we will need to ensure that table 2 is "tidy" like table 1, ensuring that each variable has its own column, each observation has its own row and each value has its own cell.
### 2. Then use the pipe (%>%) to mutate the data, calculating the rate by "(rate = cases / population * 10000)". 
### 3. Visualise the data.

```

